Contracts for Contracts

### 2023-06-14 paper/contract_solidity.tex
sec 3

We begin with specifying pre-condition and post-condition for functions
involving first-order values. Consider the following \code{toEither} example
that converts \note{Wei?} to Either:
\begin{lstlisting}[language=Solidity]
# { toEither(x) returns (y)
    requires x % 1e18 < 1e17
    ensures  y < type(uint).max / 1e18 }
function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

The first line introduces bindings for \code{toEither}'s arguments and return
values, which are \code{x} and \code{y}, respectively.
The binding names in the specification do not have to match those in the function definition.
The \code{requires}-clause specifies the pre-condition to call \code{toEither},
and the \code{ensures}-clause specifies the post-condition.
Any occurrences of \code{x} in the \code{requires}-clause refer to the actual
argument value, and similarly, occurrences of \code{y} in the \code{ensures}-clause
refer to the actual returned value.

\paragraph{Omitting}
We can omit any part of the argument or return value specification:
\begin{lstlisting}
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18 }
function toWei(uint x) returns (uint) { ... }
\end{lstlisting}
which is equivalent to core form where the omitted part is simply the boolean expression
\code{true}:
\begin{lstlisting}[language=Solidity]
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18
    requires true }
\end{lstlisting}

\paragraph{Dependent Contract}
Note that post-condition can depend on the argument as well.
\todo{example}

@@

\paragraph{Any Expression is Allowed}
When specifying the pre-/post-conditions of functions, the programmer
can use any valid Solidity expression to from the condition.
For example, Solidity has a number of built-in special variables
carrying important data (e.g. \code{msg.value}) or metadata (e.g. \code{block.timestamp})
of the transacation.
The programmer may want to check conditions against with those transactional
information.
The following code snippet shows an example that examines
\code{msg.value} as the pre-condition of a payable function:
\begin{lstlisting}
# { byTickets(n) ensures msg.value >= 1e15 * n }
function buyTickets(int n) payable { ... }
\end{lstlisting}
%\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
%\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

@@

However, addresses in Solidity have rich higher-order behaviors: they may
represent a deployed Ethereum contract and can be invoked in customized ways.
For example, the following snippet shows a function \code{f} taking an address
\code{addr} that is called at some point.
The programmer intends to the specify the precondition and postcondition when
calling \code{addr}. \note{GW: example abi?}
Nevertheless, such checks cannot be performed when \code{f} is called,
even though \code{addr} is a proper argument of \code{f}.
Because \code{addr} in this use case is effectively a function, and
it is undecidable to statically examine its properties in general.
Additionally, we may not have access to the source code of the contract
represented by \code{addr}.
\begin{lstlisting}
# { f(addr) where
    { addr(msg, amount) returns (flag, data)
      requires amount > 100
      ensures flag == true }
function f(address payable addr) {
  ... // omitting statements before the call
  (bool success, bytes memory data) =
    addr.call(abi.encodeWithSignature(
        "foo(string, uint256)", "call foo", 128
      )
    );
  ... // omitting statements after the call
}
\end{lstlisting}
\lang addresses this issue by adopting lazy checks
\cite{DBLP:conf/icfp/FindlerF02} that only take place
when the address is actually invoked \note{GW: fwd ref to impl}.
The above snippet specifies the condition of \code{addr}
in the \code{where}-clause, which requires
the \code{amount} argument must be greater than 100 and
the address call must succeed.

\paragraph{Additional Arguments}
Solidity address calls can take additional special arguments
such as \code{value} and \code{gas}:
\begin{lstlisting}
addr.call{value: msg.value, gas: 5000}(...);
\end{lstlisting}
In \lang, programmers can specify their conditions by introducing additional
bindings using the familiar syntax:
\begin{lstlisting}
# addr{value: v, gas: g}(msg, amount) returns (flag, data)
\end{lstlisting}
\code{v} and \code{g} represent the actual message value and gas
value, which can be used in \code{requires}/\code{ensures}-clauses.

\paragraph{Multiple Targets}
Multiple callee targets can inhabit in a single address value.
Programmers can specify different conditions for different
callee targets in the \code{where}-clause.
\begin{lstlisting}
TODO EXAMPLE
\end{lstlisting}

\zz{I guess we should educate the audience about the difference between low-level and high-level calls at the beginning of Section 2.4. I'll take care of this later.}

@@

\paragraph{Second-Class Enforcement}
\note{GW: an easy option is to discard enforcement of escaped addresses at all.
But it seems we can guard escaped addresses within the current contract scope.
We need to think about layered/stacked contracts of addresses (lax/picky).}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the pre-condition
and post-condition when \code{b} is called in \code{a}.

@@

Address contracts are \emph{second-class} -- we can only enforce the
pre-condition and post-condition of addresses within the current calling
context.  If the address escapes (e.g. by returning, or stored in a global
variable), we cannot enforce the contract anymore.

>>>>

We begin with specifying precondition and postcondition for functions
involving first-order values. Consider the following \code{toEither} example
that converts \note{Wei?} to Either:
\begin{lstlisting}[language=Solidity]
# { toEither(x) returns (y)
    requires x % 1e18 < 1e17
    ensures  y < type(uint).max / 1e18 }
function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

The first line introduces bindings for \code{toEither}'s arguments and return
values, which are \code{x} and \code{y}, respectively.
The binding names in the specification do not have to match those in the function definition.
The \code{requires}-clause specifies the precondition to call \code{toEither},
and the \code{ensures}-clause specifies the postcondition.
Any occurrences of \code{x} in the \code{requires}-clause refer to the actual
argument value, and similarly, occurrences of \code{y} in the \code{ensures}-clause
refer to the actual returned value.

\paragraph{Syntactic Sugars}
We can liberally omit any part of the argument or return value specification:
\begin{lstlisting}
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18 }
function toWei(uint x) returns (uint) { ... }
\end{lstlisting}
which is equivalent to core form where the omitted part is simply the boolean expression
\code{true}:
\begin{lstlisting}[language=Solidity]
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18
    requires true }
\end{lstlisting}

\paragraph{Dependent Contract}
Postconditions can depend on the argument as well. In other words,
the scope of arguments span over both the precondition and postcondition.
For example, the following snippet is a valid postcondition that requires
monotonicity.
\begin{lstlisting}
# f(x) returns (y) ensures y > x
\end{lstlisting}

@@

\paragraph{Any Expression is Allowed}
When specifying the pre-/postconditions of functions, the programmer can use
any valid Solidity expression to from the condition, including but not limited
to function calls, memory operations, and built-in special variables carrying
important transaction data (e.g. \code{msg.value}) or metadata (e.g.
\code{block.timestamp}) that are only available at runtime.
The programmers are free to check desired conditions against with these
transactional information.
For example, the following snippet examines \code{msg.value} as the
precondition of a payable function:
\begin{lstlisting}
# { byTickets(n) ensures msg.value >= 1e15 * n }
function buyTickets(int n) payable { ... }
\end{lstlisting}
With this kind of flexibility and expressiveness, programmers are able to
express invariants beyond the capability of static verification.
%\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
%\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

@@

However, addresses in Solidity have rich higher-order behaviors: they may
represent a deployed external Ethereum contract and can be invoked in
customized ways.
For example, the following snippet shows a function \code{f} taking an address
\code{addr} that is called at some point.
The programmer intends to the specify the precondition and postcondition when
calling \code{addr}. \note{GW: explain abi?}
Nevertheless, such checks against \code{addr} cannot be performed when \code{f}
is called, even though \code{addr} is a proper argument of \code{f}.
Because \code{addr} in this use case is effectively a function, and
it is undecidable to statically examine its properties in general.
Additionally, we may not have access to the source code of the contract
represented by \code{addr}.
\begin{lstlisting}
# f(addr) where
  { addr.call(msg, amount) returns (flag, data)
    requires amount > 100
    ensures flag == true }
function f(address payable addr) {
  ... // omitting statements before the call
  (bool success, bytes memory data) =
    addr.call(abi.encodeWithSignature(
        "foo(string, uint256)", "call foo", 128
      )
    );
  ... // omitting statements after the call
}
\end{lstlisting}
\lang addresses this issue by adopting lazy checks
\cite{DBLP:conf/icfp/FindlerF02} that only take place
when the address is actually invoked \note{GW: fwd ref to impl}.
The above snippet specifies the condition of \code{addr}
in the \code{where}-clause, which requires
the \code{amount} argument must be greater than 100 and
the address call must succeed.

Notably, using \lang we have separated the concern of specifications and usages
--- there is no need to change the function body or manually insert checks
around the call. This leads to more maintainable code and allows programmers to
fearlessly refactor the code.

\paragraph{Multiple Targets}
Multiple callee targets can inhabit in a single address value.
Other than the \code{call} method, there are other forms of low-level call,
e.g., \code{delegatecall}, \code{send}, etc.
% https://docs.soliditylang.org/en/v0.8.20/units-and-global-variables.html#address-related
Programmers can specify conditions for different callee targets in the
\code{where}-clause.
\begin{lstlisting}
# f(addr) where
  { addr.send(x) returns (succ) requires ... }
  { addr.call(msg, x) returns (succ, data) requires ... }
\end{lstlisting}

\zz{I guess we should educate the audience about the difference between low-level and high-level calls at the beginning of Section 2.4. I'll take care of this later.}
\note{GW: it seems we need to discern \code{call}s with different signatures}

\paragraph{Additional Arguments}
Solidity address calls can take additional special arguments
such as \code{value} and \code{gas}:
\begin{lstlisting}
addr.call{value: msg.value, gas: 5000}(...);
\end{lstlisting}
In \lang, programmers can specify their conditions by introducing additional
bindings using the familiar syntax:
\begin{lstlisting}
# addr.call{value: v, gas: g}(msg, amount) ...
\end{lstlisting}
\code{v} and \code{g} represent the actual message value and gas
value, which can be used in \code{requires}/\code{ensures}-clauses.

@@

\paragraph{Second-Class Enforcement}
\note{GW: an easy option is to discard enforcement of escaped addresses at all.
But it seems we can guard escaped addresses within the current contract scope.
We need to think about layered/stacked contracts of addresses (lax/picky).}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the precondition
and postcondition when \code{b} is called in \code{a}.

@@

Address contracts are \emph{second-class} -- we can only enforce the
precondition and postcondition of addresses within the current calling
context.  If the address escapes (e.g. by returning, or stored in a global
variable), we cannot enforce the contract anymore.
